---
title: "Derive info lesson"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#Run code here that can be accessed by all other chunks
library("learnr") 
library("tidyverse") 
# install.packages("babynames")
library("babynames") 
knitr::opts_chunk$set(echo = FALSE)
```

## Derive Information with dplyr

In this case study, you will identify the most popular American names from 1880 to 2015. While doing this, you will master three more dplyr functions:

* `mutate()`, `group_by()`, and `summarize()`, which help you use your data to compute new variables and summary statistics
  
These are some of the most useful R functions for data science, and this tutorial provides everything you need to learn them. This tutorial uses the core `tidyverse` packages, including `ggplot2`, `tibble`, and `dplyr`, and again the `babynames` package. This lesson is adapted from the RStudio lesson, Isolating Data with dplyr, as per the CC BY SA licence.  
  
### What are the most popular names of all time?

Let’s use babynames to answer a different question: what are the most popular names of all time? This question seems simple enough, but to answer it we need to be more precise: how do you define “the most popular” names? Try to think of several definitions and then click Continue.  
  
After the Continue button, you will be provided with two definitions to use.

### Most popular names

I suggest that we focus on two definitions of popular, one that uses sums and one that uses ranks:

1. **Sums** - A name is popular *if the total number of children that have the name is large when you sum across years.*
2. **Ranks** - A name is popular *if it consistently ranks among the top names from year to year.*

This raises a question:

```{r quiz-intro}
quiz(question("Do we have enough information in `babynames` to compare the popularity of names?",
    answer("No. No cell in `babynames` contains a rank value or a sum across years."),
    answer("Yes. We can use the information in `babynames` to compute the values we want.", correct = TRUE)
  ))
```

*You can use the cheunk belwo to refresh your memory about `babynames` if needed*

```{r intro, exercise=TRUE}
glimpse(babynames)
```

### Deriving information

Every data frame that you meet implies more information than it displays. For example, `babynames` does not display the total number of children who had your name, but `babynames` certainly implies what that number is. To discover the number, you only need to do a calculation:

```{r echo=TRUE, warning=FALSE}
babynames %>% 
  filter(name == "Garrett", sex == "M") %>% 
  summarise(total = sum(n))
```

### Useful functions

`dplyr` provides three functions that can help you reveal the information implied by your data:

* `summarise()`
* `group_by()`
* `mutate()`

Like `select()`, `filter()` and `arrange()`, these functions all take a data frame as their first argument and return a new data frame as their output, which makes them easy to use in pipes.  
  
Let’s master each function and use them to analyse popularity as we go.  

## summarise()

`summarise()` takes a data frame and uses it to calculate a new data frame of summary statistics.

### Syntax

To use `summarise()`, pass it a data frame and then one or more named arguments. Each named argument should be set to an R expression that generates a single value. Summarise will turn each named argument into a column in the new data frame. The name of each argument will become the column name, and the value returned by the argument will become the column contents.

### Example

Here we can use `summarise()` from above to calculate the total number of boys named “Garrett”, but let’s expand that code to also calculate

* `max` - the maximum number of boys named “Garrett” in a single year
* `mean` - the mean number of boys named “Garrett” per year
* `sd` - the standard deviation of boys named “Garrett” across years

<div class="alert alert-warning">
**TIP**: By default, all R functions operating on vectors that contains **missing data** will return `NA`. It’s a way to make sure that users know they have **missing data**, and make a conscious decision on how to deal with it. When dealing with simple statistics like the `mean`, the easiest way to ignore `NA` (the missing data) is to use `na.rm = TRUE` (rm stands for remove). See the below code. 
</div>

```{r echo=TRUE}
babynames %>% 
  filter(name == "Garrett", sex == "M") %>% 
  summarise(total = sum(n, na.rm = TRUE), 
            max = max(n, na.rm = TRUE), 
            mean = mean(n, na.rm = TRUE),
            sd = sd(n, na.rm = TRUE))
```

Don’t let the code above fool you. The first argument of `summarise()` is always a data frame, but when you use `summarise()` in a pipe, the first argument is provided by the pipe operator, `%>%`. Here the first argument will be the data frame that is returned by  `babynames %>% filter(name == "Garrett", sex == "M")`.  

<div class="alert alert-success">
**Question:** How would you describe in words the process that is being undertaken in the above code. Remember that `%>%` can the refereed to as "then".
</div>

### Exercise - summarise()

*Use the code chunk below to compute three statistics:*

* the total number of children who ever had your name
* the mean number of children given your name per year
* the standard error of the mean (remember back to the workshop this week)
  
If you cannot think of an R function that would compute each statistic, click the Hint/Solution button.

```{r summ2, exercise=TRUE}

```

```{r summ2-hint-1}
#To calculate the number of replicates
reps = sum(!is.na(n))
```

```{r summ2-hint-2}
#SE is the standard deviation divided by the square root of the number of replicates
se = sd(n, na.rm = TRUE)/sqrt(reps)
```

```{r summ2-hint-3}
babynames %>% 
  filter(name == "Beyonce", sex == "F") %>% 
  summarise(total = sum(n, na.rm = TRUE),  
            mean = mean(n, na.rm = TRUE),
            reps = sum(!is.na(n)),
            se = sd(n, na.rm = TRUE)/sqrt(reps))
```

<div class="alert alert-success">
**QUESTION:** Is the estimate for the SE of the mean informing you about the spread of the data (i.e. the variability in how many children were named your name each year), or the precision in which you could estimate the average number of children given your name per year?.
</div>

## Summary functions

So far our `summarise()` examples have relied on `sum()`, `max()`, and `mean()`. But you can use any function in `summarise()` so long as it meets one criteria: the function must take a vector of values as input and return a single value as output. Functions that do this are known as summary functions and they are common in the field of descriptive statistics. Some of the most useful summary functions include:  

* **Measures of location** - `mean(x)`, `median(x)`, `quantile(x, 0.25)`, `min(x)`, and `max(x)`
* **Measures of spread** - `sd(x)`, `var(x)`, `IQR(x)`, and `mad(x)`
* **Measures of position** - `first(x)`, `nth(x, 2)`, and `last(x)`
* **Counts** - `n_distinct(x)` and `n()`, which takes no arguments, and returns the size of the current group or data frame.
* **Counts of logical values** - `sum(!is.na(x)`, which counts the number of TRUEs returned by a logical test.  

<div class="alert alert-warning">
**TIP**: I recommend whenever you are trying to count the number of replicates or rows in a sample that you use `sum(!is.na(x)`. This will ignore all **missing data**, indicated by `NA`. 
</div>
  
Let’s apply some of these summary functions. Click Continue to test your understanding.  

### Khaleesi challenge
  
“Khaleesi” is a very modern name that appears to be based on the Game of Thrones TV series, which premiered on April 17, 2011. What year did the Khaleesi name first arise in the `babynames` data? In the chunk below, filter babynames to just the rows where name == “Khaleesi”. Then use `summarise()` and a summary function (use those from above) to return the first value of `year` in the data set.

```{r khaleesi, exercise=TRUE}


```

```{r khaleesi-hint}
babynames %>% 
  filter(name == "Khaleesi") %>% 
  summarise(year = first(year))
```

### Distinct name challenge

In the chunk below, use `summarise()` and a summary function to return a data frame with two columns:

* A column named `n` that displays the total number of names for which there is data across all years in `babynames`
* A column named `distinct` that displays the number of distinct names in `babynames`

Will these numbers be different? Why or why not?

```{r names, exercise=TRUE}

```

```{r names-hint}
babynames %>% 
  summarise(n = sum(!is.na(name)), distinct = n_distinct(name))
```

### summarise by groups?

How can we apply `summarise()` to find the most popular names in `babynames`? You’ve seen how to calculate the total number of children that have your name, which provides one of our measures of popularity, i.e. the total number of children that have a name:  

```
babynames %>% 
  filter(name == "Beyonce", sex == "F") %>% 
  summarise(total = sum(n))
```

However, we had to isolate your name from the rest of your data to calculate this number. You could imagine writing a program that goes through each name one at a time and:

1. filters out the rows with just that name
2. applies summarise to the rows  
  
Eventually, the program could combine all of the results back into a single data set. However, you don’t need to write such a program; this is the job of dplyr’s `group_by()` function.

## group_by()

`group_by()` takes a data frame and then the names of one or more columns in the data frame. It returns a copy of the data frame that has been “grouped” into sets of rows that share identical combinations of values in the specified columns.  

### group_by() in action
  
By itself, `group_by()` doesn’t do much. It assigns grouping criteria that is stored as metadata alongside the original data set. If your dataset is a tibble, as above, R will tell you that the data is grouped at the top of the tibble display. In all other aspects, the data looks the same.

However, when you apply a dplyr function like `summarise()` to grouped data, dplyr will execute the function in a groupwise manner. Instead of computing a single summary for the entire data set, dplyr will compute individual summaries for each group and return them as a single data frame. The data frame will contain the summary columns as well as the columns in the grouping criteria, which makes the result decipherable:

```{r group1, exercise=TRUE, exercise.eval=TRUE}
babynames %>%
  group_by(year, sex) %>% 
  summarise(total = sum(n))
```

To understand exactly what `group_by()` is doing, remove the line `group_by(year, sex) %>%` from the code above and rerun it. How do the results change?

### Sometimes you need to un_group()

After grouping data with `group_by()`, there may sometimes be a need to return to a non-grouped form. This can happen if you need to apply a function to each group, such as standardising your data by sex, and then wish to keep working with the data as a whole. If you forgot to ungroup first, then any subsequent applied functions will still be acting on a per group basis. If you wish to manually remove the grouping criteria from a data set, you can do so with `ungroup()`.
```
babynames %>%
  group_by(year, sex) %>% 
  ungroup()
```

### The most popular names by total children
  
You now know enough to calculate the most popular names since 1880 by total children (it may take some strategising, but you can do it!).

In the code chunk below, use group_by(), summarise(), and arrange() to display the most popular names. Compute popularity as the total number of children of a single gender given a name. In other words, the total number of boys named “Kelly” should be computed separately from the total number of girls named “Kelly”.  

```{r pop, exercise=TRUE}

```

```{r pop-hint}
babynames %>%
  group_by(name, sex) %>% 
  summarise(total = sum(n)) %>% 
  arrange(desc(total))
```
  
### The history of the most popular names by total children

Let’s examine how the popularity of popular names has changed over time. To help us, I have plotted the births over time for the 10 most popular names from above. Do you think using the total number of children born is the best predictor of a names popularity? How would the rapid population growth that occured after 1920 bias the results? Why are so few girls names present in the top 10?

```{r}
top10 <- babynames %>%
  group_by(name, sex) %>% 
  summarise(total = sum(n)) %>% ungroup %>%
  top_n(10, total)

top_10_comp <- babynames %>% semi_join(top10, by = c("name", "sex"))

top_10_comp %>% 
  ggplot() + 
  labs(title = "Most popular names based on total numbers") +
  ylab("Births per year") + xlab("Year of birth") +
  scale_x_continuous(breaks = seq(1880, 2025, by = 10), limits = c(1880, 2020)) +
  geom_line(aes(x = year, y = n, color = name), na.rm = TRUE) +
  theme_bw() + theme(legend.title=element_blank())
```

## mutate()

Our first definition of popularity is confounded with population growth. If more children are born each year, the number of children who receive a name could grow even if the proportion of children given that name declines. The most popular names in 2015 likely represent far more children than the most popular names in 1880. The total number of children given a name may still be the best definition of popularity to use, but it will overweight names that have been popular in recent years. There is also evidence that our definition is confounded with a gender effect: only one of the top ten names was a girl’s name.

### Popularity based on rank

If you are concerned about these things, you might prefer to use our second definition of popularity, which would give equal representation to each year and gender: **Ranks** - A name is popular if it consistently ranks among the top names from year to year.  
  
To use this definition, we could:
  
1. Compute the rank of each name within each year and gender. The most popular name would receive the rank 1 and so on.
2. Find the median rank for each name, accounting for gender. The names with the lowest median would be the names that “consistently rank among the top names from year to year.”  
  
To do this, we will need to learn one last dplyr function.  

### mutate()

`mutate()` uses a data frame to compute new variables. It then returns a copy of the data frame that includes the new variables. For example, we can use `mutate()` to compute a percent variable for babynames. Here percent is just the prop multiplied by 100 and rounded to two decimal places.

```{r echo=TRUE}
babynames %>%
  mutate(percent = round(prop * 100, 2))
```

### An exercise in mutation

The syntax of mutate is similar to `summarise()`. `mutate()` takes first a data frame, and then one or more named arguments that are set equal to R expressions. `mutate()` turns each named argument into a column. The name of the argument becomes the column name and the result of the R expression becomes the column contents.  
  
Like `summarise()`, `mutate()` works in combination with a specific type of function. `summarise()` expects summary functions, which take a vector of input and return a single value. `mutate()` expects vectorized functions, which take a vector of input and return a vector of values.  
  
In other words, summary functions like `min()` and `max()` won’t work well with `mutate()`. You can see why if you take a moment to think about what `mutate()` does: `mutate()` adds a new column to the original data set. In R, every column in a dataset must be the same length, so `mutate()` must supply as many values for the new column as there are in the existing columns.  
  
If you give `mutate()` an expression that returns a single value, it will follow R’s recycling rules and repeat that value as many times as needed to fill the column. This can make sense in some cases, but the reverse is never true: you cannot give `summarise()` a vectorized function; `summarise()` needs its input to return a single value.  
  
What are some of R’s vectorized functions? Click Continue to find out.  

### The most useful vectorized functions

Some of the most useful vectorised functions in R to use with mutate() include:  

1. **Arithmetic operators**: `+`, `-`, `*`, `/`, `^`. These are all vectorised, using R’s so called “recycling rules”. If one vector of input is shorter than the other, it will automatically be repeated multiple times to create a vector of the same length.
2. **Modular arithmetic**: `%/%` (integer division) and `%%` (remainder)
3. **Logical comparisons**: `<`, `<=`, `>`, `>=`, `!=`
4. **Logs**: `log(x)`, `log2(x)`, `log10(x)`
5. **Offsets**: `lead(x)`, `lag(x)`
6. **Cumulative aggregates**: `cumsum(x)`, `cumprod(x)`, `cummin(x)`, `cummax(x)`, `cummean(x)`
7. **Ranking**: `min_rank(x)`, `row_number(x)`, `dense_rank(x)`, `percent_rank(x)`, `cume_dist(x)`, `ntile(x)`  
  
For ranking, I recommend that you use `min_rank()`, which gives the smallest values the top ranks. To rank in descending order, use the familiar `desc()` function, e.g.

```{r echo=TRUE}
min_rank(c(50, 100, 1000))
```

```{r echo=TRUE}
min_rank(desc(c(50, 100, 1000)))
```

### An exercise in ranking
  
Let’s practice by ranking the entire dataset based on prop. In the chunk below, use mutate() and min_rank() to creat a new column that ranks each row based on its prop value, with the highest values receiving the top ranks.

```{r mut,  exercise=TRUE}

```

```{r mut-hint}
babynames %>% 
  mutate(rank = min_rank(desc(prop)))
```

### Rankings by group

In the previous exercise, we assigned rankings across the entire data set. For example, with the exception of ties, there was only one 1 in the entire data set, only one 2, and so on. To calculate a popularity score across years, you will need to do something different: you will need to assign rankings within groups of year and sex.  
  
To rank within groups, combine `mutate()` with `group_by()`. Like dplyr’s other functions, `mutate()` will treat grouped data in a group-wise fashion.

Add `group_by()` to our code from above, to calculate ranking within year and sex combinations. Do you notice the numbers change?

```{r mut2,  exercise=TRUE}
babynames %>% 
  mutate(rank = min_rank(desc(prop)))
```

```{r mut2-hint}
babynames %>% 
  group_by(year, sex) %>% 
  mutate(rank = min_rank(desc(prop)))
```

### The most popular names by yearly rankings

`group_by()` provides the missing piece for calculating our second measure of popularity. In the code chunk below,  

* Group `babynames` by `year` and `sex`
* Assign a rank to each name based on descending values of prop
* Regroup the data by `name` and `sex`
* Compute the median ranking for each name and sex combination
* Arrange the results so the names with the lowest sum appear at the top of the data set.

```{r mut3, exercise=TRUE}

```

```{r mut3-hint}
babynames %>% 
  group_by(year, sex) %>% 
  mutate(rank = min_rank(desc(prop))) %>% 
  group_by(name, sex) %>% 
  summarise(score = median(rank)) %>% 
  arrange(score)
```

### Recap

Here is the plot of births over time for the 10 most popular names based on rank alone.

```{r echo=FALSE}
top10rank <- babynames %>% 
  group_by(year, sex) %>% 
  mutate(rank = min_rank(desc(prop))) %>% 
  group_by(name, sex) %>% 
  summarise(score = median(rank)) %>% 
  arrange(score) %>% ungroup %>%
  top_n(10, desc(score))

top_10_rank_c <- babynames %>% semi_join(top10rank, by = c("name", "sex"))

top_10_rank_c %>% 
  ggplot() + 
  labs(title = "Most popular names based on rank") +
  ylab("Births per year") + xlab("Year of birth") +
  scale_x_continuous(breaks = seq(1880, 2025, by = 10), limits = c(1880, 2020)) +
  geom_line(aes(x = year, y = n, color = name), na.rm = TRUE) +
  theme_bw() + theme(legend.title=element_blank())
```

In this lesson, you learned three functions for isolating data within a table:  

* `select()`
* `filter()`
* `arrange()`  
  
You also learned three functions for deriving new data from a table:  
  
* `summarise()`
* `group_by()`
* `mutate()`  
  
Together these six functions create a grammar of data manipulation, a system of verbs that you can use to manipulate data in a sophisticated, step-by-step way. These verbs target the everyday tasks of data analysis. No matter which types of data you work with, you will discover that:  
  
1. Data sets often contain more information than you need
2. Data sets imply more information than they display  
  
The six dplyr functions help you work with these realities by isolating and revealing the information contained in your data. In fact, dplyr provides more than six functions for this grammar: dplyr comes with several functions that are variations on the themes of select(), `filter()`, `summarise()`, and `mutate()`. Each follows the same pipeable syntax that is used throughout dplyr. If you are interested, you can learn more about these peripheral functions in the dplyr cheatsheet handed out today.  
  
Now on to the final exercise of the day.
