---
title: "Week 1 Practical"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#Run code here that can be accessed by all other chunks
rm(list = ls())
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Installing packages

R is an open-source programming language, meaning that users can contribute packages that make our lives easier, and we can use them for free. In R, the fundamental unit of shareable code is the package. A package bundles together code, data, documentation, and tests, and is easy to share with others. As of January 2015, there were over 6,000 packages available on the Comprehensive R Archive Network, or CRAN, the public clearing house for R packages. This huge variety of packages is one of the reasons that R is so successful: the chances are that someone has already solved a problem that you’re working on, and you can benefit from their work by downloading their package.   
  
You install them from CRAN with `install.packages("x")`.  
You use them in R with `library(x)`.  
You get help on them with `package?x` and `help(package = "x")`.  

For this lab, and many others in the future, we will use the following R packages:

* `dplyr`: for data wrangling  
* `gpglot`: for data visualisation  

The easiest way to load these packages, any many others that become increasingly useful as you get better at R, is to load the `tidyverse` package. 

### Exercise in loading packages

*Run the following code to load the tidyverse package and notice the resulting output. Conflicts arise because some packages can use the same name for two different functions. Here `dply` is using `filter`, as is the bases `stats` package. Because tidyverse was loaded last, it takes priority.*

```{r install, exercise=TRUE}
library(tidyverse)
```

```{r quiz1}
quiz(
  question("How many packages load as part of `tidyverse`",
    answer("0"),
    answer("1"),
    answer("3"),
    answer("5"),
    answer("6", correct = TRUE)
  ))
```

### Exercise in finding help

*Use the code chunk below to get help on the package `readr`.*

```{r help, exercise=TRUE}

```

```{r help-hint-1}
package?...
```

```{r help-hint-2}
package?readr
```

```{r quiz2}
quiz(
  question("What is the purpose of the `readr` package?",
    answer("To provide a complete and consistent functional programming toolkit for R"),
    answer("To provide a fast and friendly way to read rectangular data", correct = TRUE),
    answer("To provide a flexible grammar of data manipulation"),
    answer("To provide a system for 'declaratively' creating graphic")
  )
)
```

### Exercise in finding help again

A code chunk to use and some questions to verify that you understand the purposes of various base and recommended R packages.

```{r quizz1, exercise=TRUE}

```

```{r quiz3}
quiz(
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("ggplot2", correct = TRUE)
  )
)
```

## Basic R coding



### Exercise in programming

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*


```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:
