---
title: "Isolate data lesson"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#PLEASE MAKE SURE ALL PACKAGES ARE INSTALLED FIRST

#Run code here that can be accessed by all other chunks
library("learnr") 
library("tidyverse") 
library("babynames") 
knitr::opts_chunk$set(echo = FALSE)
```

## Isolating Data with dplyr

In this case study, you will explore the popularity of your own name over time. Along the way, you will master some of the most useful functions for isolating variables, cases, and values within a data frame:

* `select()` and `filter()`, which let you extract rows and columns from a data frame  
* `arrange()`, which lets yuo reorder the rows in your data  
* `%>%`, which organizes your code into reader-friendly “pipes”  

This tutorial uses the core tidyverse packages, including `ggplot2`, `tibble`, and `dplyr`, as well as the `babynames` package. The `babynames` package contains all names used for at least 5 children of either sex in the US from 1880 to 2015. This lesson is adapted from the RStudio lesson, Isolating Data with dplyr, as per the CC BY SA licence. 

Click the Next Topic button to begin.
  
## The history of your name

You can use the data in babynames to make graphs like this, which reveal the history of a name, perhaps your name. But before you do, you will need to trim down `babynames`. At the moment, there are more rows in `babynames` than you need to build your plot. Your goal in this section is to repeat this process for your own name (or a name that you choose). Along the way, you will learn a set of functions that isolate information within a data set.  

```{r echo=FALSE, warning=FALSE}
baby_edit <- babynames %>% 
  filter((name == "Matthew" & sex == "M") |
           (name == "Stephanie" & sex == "F") |
           (name == "Evatt" & sex == "M") | 
           (name == "Tom" & sex == "M") |
           (name == "Marcus" & sex == "M")) %>% 
  select(name, year, prop, n) 
  
evatt_add <-  tibble(name = "Evatt", year =  unique(baby_edit$year), prop = 0, n = 0)

baby_edit <- baby_edit %>% bind_rows(evatt_add)

ggplot(baby_edit, aes(x = year, y = prop, colour = name)) +
  geom_line(na.rm = TRUE) + 
  labs(title = "Popularity of names for your teaching team") +
  ylab("Proprtion of all names") + xlab("Year of birth") +
  scale_x_continuous(breaks = seq(1910, 2025, by = 10), limits = c(1910, 2020)) +
  theme_bw() + 
  theme(legend.justification=c(0,1), 
        legend.position=c(0,1),
        legend.background = element_blank(),
        legend.title=element_blank())
```
  
To see what I mean, consider the size and complexity of the `babynames` data. There are `r length(unique(babynames$name))` unique baby names spanning the years 1880 to 2015. You will be working with a data set that is 1,858,689 rows long. Isolating the data that you need is crucial.

```{r str, exercise=TRUE, exercise.eval=FALSE}
glimpse(babynames)

```

### Isolating data

This type of task occurs often in Data Science: you need to extract data from a table before you can use it. You can do this task quickly with three functions that come in the dplyr package:

* `select()` - which extracts columns from a data frame  
* `filter()` - which extracts rows from a data frame  
* `arrange()` - which moves important rows to the top of a data frame  
  
Each function takes a data frame or tibble as it’s first argument and returns a new data frame or tibble as its output.

## select()
  
`select()` extracts columns of a data frame and returns the columns as a new data frame. To use `select()`, pass it the name of a data frame to extract columns from, and then the names of the columns to extract. The column names do not need to appear in quotation marks or be prefixed with a `$`; `select()` knows to find them in the data frame that you supply.
  
*Use the example below to get a feel for `select()`. Can you extract just the name column? How about the name and year columns? How about all of the columns except prop?*

```{r select, exercise=TRUE, exercise.eval=TRUE}
select(babynames, name, sex)


```

```{r select-hint}
select(babynames, name)
select(babynames, name, year)
select(babynames, year, sex, name, n)
```

### select() helpers
  
*You can also use a series of helpers with `select()`. For example, if you place a minus sign before a column name,  `select()` will return every column but that column. Can you predict how the minus sign will work here?*

```{r select2, exercise=TRUE}
select(babynames, -c(n, prop))


```

The table below summarizes the other select() helpers that are available in dplyr. Study it, and then click “Continue” to test your understanding.

Helper Function	|  Use                                         | Example
:---------------|:---------------------------------------------|:--------
`-`	            | Columns except                               |	`select(babynames, -prop)`
`:`	            | Columns between (inclusive)                  | `select(babynames, year:n)`
`contains()`    |	Columns that contains a string               | `select(babynames, contains("n"))`
`ends_with()`   | Columns that ends with a string              | `select(babynames, ends_with("n"))`
`matches()`	    | Columns that matches a regex                 | `select(babynames, matches("n"))`
`num_range()`	  | Columns with a numerical suffix in the range |	Not applicable with babynames
`one_of()`     	| Columns whose name appear in the given set   |	`select(babynames, one_of(c("sex", "gender"))`
`starts_with()`	| Columns that starts with a string	           | `select(babynames, starts_with("n"))`
  
```{r quiz-select}
quiz(caption = "Quiz: You can use the code chunk above to help answer this question.",
  question("Which of these is not a way to select the name and n columns together?",
    answer("select(babynames, -c(year, sex, prop))"),
    answer("select(babynames, name:n)"),
    answer('select(babynames, starts_with("n"))'),
    answer('select(babynames, ends_with("n"))', correct = TRUE)
  ))
```

## filter()

`filter()` extracts rows from a data frame and returns them as a new data frame. As with `select()`, the first argument of `filter()` should be a data frame to extract rows from. The arguments that follow should be logical tests; `filter()` will return every row for which the tests return `TRUE`.
  
### filter in action 
  
For example, the code chunk below returns every row with the name “Sea” in babynames.

```{r filter1, echo=TRUE}
filter(babynames, name == "Sea")
```
  
### Logical tests
  
To get the most from filter, you will need to know how to use R’s logical test operators, which are summarised below.  
  
Logical operator	| Tests                                         | Example
:-----------------|:----------------------------------------------|:-------
**>**	            | Is `x` greater than `y`?                      | `x > y` 
**>=**	          | Is `x` greater than or equal to `y`?          | `x >= y` 
**<**	            | Is `x` less than `y`?	                        | `x < y`
**<=**            | Is `x` less than or equal to `y`?	            | `x <= y`
**==**	          | Is `x` equal to `y`?                          | `x == y`
**!=**            | Is `x` not equal to `y`?                      | `x != y`
**is.na()**       | Is `x` an `NA`?	                              | `is.na(x)`
**!is.na()**      | Is `x` not an `NA`?                           | `!is.na(x)`

###Exercise - Logical Operators

See if you can use the logical operators to manipulate our code below to show:

* All of the names where prop is greater than or equal to 0.08
* All of the children named “Khaleesi”
* All of the names that have a missing value for `n` (Hint: this should return an empty data set).

```{r filter2, exercise=TRUE}
filter(babynames, name == "Sea")


```

```{r filter2-hint}
filter(babynames, prop >= 0.08)
filter(babynames, name == "Khaleesi")
filter(babynames, is.na(n))
```

### Two common mistakes
  
When you use logical tests, be sure to look out for two common mistakes. One appears in each code chunk below. Can you find them? When you spot a mistake, fix it and then run the chunk to confirm that it works.

```
filter(babynames, name = "Sea")
```
```{r filter3, exercise=TRUE}

```

```{r filter3-hint-1}
filter(babynames, name == "Sea")
```

```
filter(babynames, name == Sea)
```
```{r filter4, exercise=TRUE}

```

```{r filter4-hint-1}
filter(babynames, name == "Sea")
```

### Two mistakes - Recap

When you use logical tests, be sure to look out for these two common mistakes:

1. using `=` instead of `==` to test for equality.
2. forgetting to use quotation marks when comparing strings, e.g. `name == Abby`, instead of `name == "Abby"`

### Combining tests

If you provide more than one test to `filter()`, `filter()` will combine the tests with an and statement (`&`): it will only return the rows that satisfy all of the tests.

To combine multiple tests in a different way, use R’s Boolean operators. For example, the code below will return all of the children named Sea or Anemone.

```{r filter5, echo = TRUE}
filter(babynames, name == "Sea" | name == "Anemone")
```

### Boolean operators

You can find a complete list or base R’s boolean operators in the table below.

Boolean operator | Represents | Example
:----------------|:-----------|:---------
**&** | Are *both* `A` and `B` true? | `A & B`
**\|**	| Are *one or both* of `A` and `B` true? | `A | B`
**!**	| Is `A` *not* true?	| `!A`
**xor()**	| Is *one and only one* of `A` and `B` true?	| `xor(A, B)`
**%in%**	| Is x in the set of `A`, `B`, and `C`?	| `x %in% c(A, B, C)`
**any()**	| Are *any* of `A`, `B`, or `C` true?	| `any(A, B, C)`
**all()**	| Are *all* of `A`, `B`, or `C` true?	| `all(A, B, C)`

### Exercise - Combining tests

Use Boolean operators to alter the code chunk below to return only the rows that contain:

* *Girls* named Sea
* Names that were used by exactly 5 or 6 children in 1880
* Names that are one of Acura, Lexus, or Yugo

```{r filter6, exercise=TRUE, exercise.eval=TRUE}
filter(babynames, name == "Sea" | name == "Anemone")
```

```{r filter6-hint}
filter(babynames, name == "Sea", sex == "F")
filter(babynames, n == 5 | n == 6, year == 1880)
filter(babynames, name %in% c("Acura", "Lexus", "Yugo"))
```

### Two more common mistakes

Logical tests also invite two common mistakes that you should look out for. Each is displayed in a code chunk below, one produces an error and the other is needlessly verbose. Diagnose the chunks and then fix the code (reference Boolean operators above).

```
filter(babynames, 10 < n < 20)
```
```{r filter7, exercise=TRUE}

```

```{r filter7-hint}
filter(babynames, n > 10, n < 20)
# or
filter(babynames, 10 < n, n < 20)
```

```
filter(babynames, n == 5 | n == 6 | n == 7 | n == 8 | n == 9)
```
```{r filter8, exercise=TRUE}

```

```{r filter8-hint}
filter(babynames, n %in% c(5, 6, 7, 8, 9))
```

### Two more common mistakes - Recap

When you combine multiple logical tests, be sure to look out for these two common mistakes:

1. Collapsing multiple logical tests into a single test without using a boolean operator
2. Using repeated `|` instead of `%in%`, e.g. `x == 1 | x == 2 | x == 3` instead of `x %in% c(1, 2, 3)`  
  
## arrange()

`arrange()` returns all of the rows of a data frame reordered by the values of a column. As with `select()`, the first argument of `arrange()` should be a data frame and the remaining arguments should be the names of columns. If you give `arrange()` a single column name, it will return the rows of the data frame reordered so that the row with the lowest value in that column appears first, the row with the second lowest value appears second, and so on. If the column contains character strings, `arrange()` will place them in alphabetical order.

### Exercise - arrange()

Use the code chunk below to arrange babynames by `n`. Can you tell what the smallest value of `n` is?

```{r arrange1, exercise=TRUE}

```

```{r arrange1-hint}

```

### Tie breakers

If you supply additional column names, `arrange()` will use them as tie breakers to order rows that have identical values in the earlier columns. Add to the code below, to make `prop` a tie breaker. The result should first order rows by value of `n` and then reorder rows within each value of `n` by values of `prop`.

```{r arrange2, exercise=TRUE, exercise.eval=TRUE}
arrange(babynames, n)
```

```{r arrange2-hint, exercise=TRUE}
arrange(babynames, n, prop)
```

### desc()

If you would rather arrange rows in the opposite order, i.e. from large values to small values, surround a column name with `desc()`. `arrange()` will reorder the rows based on the largest values to the smallest.

Add a `desc()` to the code below to display the most popular name for 2015 (the largest year in the dataset) instead of 1880 (the smallest year in the dataset).

```{r arrange3, exercise=TRUE, exercise.eval=TRUE}
arrange(babynames, year, desc(prop))
```

```{r arrange3-hint}
arrange(babynames, desc(year), desc(prop))
```

Think you have it? Click Continue to test yourself.

### arrange() quiz

Which name was the most popular for a single gender in a single year? In the code chunk below, use arrange() to make the row with the largest value of prop appear at the top of the data set.

```{r arrange4, exercise=TRUE}

```

```{r arrange4-hint}
arrange(babynames, desc(prop))
```

Now arrange babynames so that the row with the largest value of n appears at the top of the data frame. Will this be the same row? Why or why not?

```{r arrange5, exercise=TRUE}

```

```{r arrange5-hint}
arrange(babynames, desc(n))
# The number of children represented by each proportion grew 
# over time as the population grew
```

## %>%

### Steps

Notice how each dplyr function takes a data frame as input and returns a data frame as output. The dplyr functions never modify their inputs, so if you want to save the result, you’ll need to use the assignment operator, `<-`. This makes the functions easy to use in a step by step fashion. For example, you could:

```{r steps, echo=TRUE}
boys_2015 <- filter(babynames, year == 2015, sex == "M")
boys_2015 <- select(boys_2015, name, n)
boys_2015 <- arrange(boys_2015, desc(n))
boys_2015
```

### Redundancy

The result shows us the most popular boys names from 2015, which is the most recent year in the data set. But take a look at the code. Do you notice how we re-create `boys_2015` at each step so we will have something to pass to the next step? This is an inefficient way to write R code.

You could avoid creating `boys_2015` by nesting your functions inside of each other, but this creates code that is hard to read:

```
arrange(select(filter(babynames, year == 2015, sex == "M"), name, n), desc(n))
```

The dplyr package provides a third way to write sequences of functions: the pipe.
  
### The pipe

The pipe operator `%>%` performs an extremely simple task: it passes the result on its left into the first argument of the function on its right. Or put another way, `x %>% f(y)` is the same as `f(x, y)`. This piece of code punctuation makes it easy to write and read series of functions that are applied in a step by step way. For example, we can use the pipe to rewrite our code above:

```{r pipe1, echo = TRUE}
babynames %>% 
  filter(year == 2015, sex == "M") %>% 
  select(name, n) %>% 
  arrange(desc(n))
```

As you read the code, pronounce `%>%` as “then”. You’ll notice that dplyr makes it easy to read pipes. Each function name is a verb, so our code resembles the statement, “Take babynames, then filter it by name and sex, then select the name and n columns, then arrange the results by descending values of n.”

dplyr also makes it easy to write pipes. Each dplyr function returns a data frame that can be piped into another dplyr function, which will accept the data frame as its first argument. In fact, dplyr functions are written with pipes in mind: each function does one simple task. dplyr expects you to use pipes to combine these simple tasks to produce sophisticated results.
  
### Exercise - Pipes

We will use pipes for the remainder of the tutorial, and I will expect you to as well. Let’s practice a little by writing a new pipe in the chunk below. The pipe should:

1. Filter babynames to just the girls that were born in 2015
2. Select the name and n columns
3. Arrange the results so that the most popular names are near the top.

Try to write your pipe without copying and pasting the code from above.

```{r pipe2, exercise=TRUE}

```

```{r pipe2-hint}
babynames %>% 
  filter(year == 2015, sex == "F") %>% 
  select(name, n) %>% 
  arrange(desc(n))
```
  
### Your name

You’ve now mastered a set of skills that will let you easily plot the popularity of your name over time. In the code chunk below, use a combination of dplyr and ggplot2 functions with `%>%` to:

1. Trim `babynames` to just the rows that contain your name and your sex
2. Trim the result to just the columns that will appear in your graph (not strictly necessary, but useful practice)
3. Assign the results of these steps to the variable `my_name`
3. Plot the results of `my_name` as a line graph with `year` on the x axis and `n` on the y axis (this code is provided; next week we cover ggplot)
  
```{r pipe3, exercise=TRUE}
#Enter your code below here:




#Here is the ggplot code provided. 
my_name %>% ggplot(aes(x = year, y = prop)) +
  geom_line(na.rm = TRUE) +
  labs(title = "Popularity of your name") +
  ylab("Proprtion of all names") + xlab("Year of birth") +
  scale_x_continuous(breaks = seq(1910, 2025, by = 10), limits = c(1910, 2020)) +
  theme_bw()
```

```{r pipe3-hint-1}
my_name <- ...
```

```{r pipe3-hint-2}
my_name <- babynames %>% 
  ...
```

```{r pipe3-hint-3}
my_name <- babynames %>% 
  filter(name == "Matthew", sex == "M") %>% 
  select(year, n, prop)
```

### Recap

Together, `select()`, `filter()`, and `arrange()` let you quickly find information displayed within your data. The next lesson will show you how to derive information that is implied by your data, but not displayed within your data set. In that tutorial, you will continue to use the `%>%` operator, which is an essential part of programming in the `tidyverse` world and now BIO3011. Pipes help make R expressive, like a spoken language. Spoken languages consist of simple words that you combine into sentences to create sophisticated thoughts. In the `tidyverse`, functions are like words: each does one simple task well. You can combine these tasks into pipes with `%>%` to perform complex, customized procedures.
